### 2020-12-04 doc/manuscript/Dow_manuscript.Rmd
some work on methods, auto fill some statistics

#Table numbers
hypothesis_table=1
species_table=2
mean_growth_parameters_table=3

@@

@@

Metal dendrometer bands installed on *99-145* and *700-755* trees within the SCBI and Harvard Forest ForestGEO plots, respectively, were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 and 1998-2003, respectively. The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census (Table `r sample_size_by_year`) *(tags_per_year_preclean in data folder)*. Across years, the number of bands sampled averaged *~135* at SCBI and  *~717* at Harvard Forest. At SCBI, four species were sampled including two diffuse-porous (Lirodendron tulipfera and Fagus grandafolia) and two ring-porous (Quercus rubra and Quercus alba) species. At Harvard Forest, eleven species were sampled including eight diffuse-porous and three ring-porous (Table `r species_table`).

```{r species_table, eval=TRUE, echo=FALSE, warning=FALSE}
knitr::include_graphics("tables_figures/species_table.png")
```

@@

### Tree-rings
Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
We analyzed records for the dominant ring- and diffuse- porous species at each site (Table `r species_table`), excluding species with other xylem architectures (*Juglans nigra* L. at SCBI, *Tsuga Canadensis* (L.) Carrière at Harvard Forest)
At SCBI,  were collected as described in @helcoski_growing_2019. 
Briefly, ...  (*summarize / reword from Helcoski*). 
In total, we analyzed # cores representing eleven of the twelve species contributing most to aboveground woody productivity within the SCBI ForestGEO plot.
At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
In total, we analyzed # cores from three species.

@@

### Climate
Daily maximum temperatures ($T_{MAX}$) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot (data can be found on the ForestGEO climate portal, DOI on github). Daily maximum temperatures at Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).  The R package climpact () was used to plot $T_{MAX}$ for visual inspection. Climpact also identifies readings that were greater than 3 standard deviations away from yearly means. These points were labeled as outliers and removed from the dataset. Gaps in the SCBI meteorological tower data were subsequently filled using $T_{MAX}$ readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail). (*how were gaps filled for Harvard? Or was the record good?*)

(*climate data for tree rings-- KAT can write this*)

@@

```{r climwin, out.width="100%", echo=FALSE, fig.align="center", fig.cap= climwin_caption}
# Created in RScripts/additional_figures.R
knitr::include_graphics("tables_figures/climwin_figure.png")
```

@@

```{r tree_rings, out.width="100%", echo=FALSE, fig.align="center", fig.cap= tree_rings_caption}
# Created in RScripts/additional_figures.R
knitr::include_graphics("tables_figures/tree_cores_tmax.png")
```

@@

## (summary of key findings, gap filled)
We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity...

## ("limitation in filling gap, how to interpret/fix")
Although the number of years analyzed here is limited (# years at SCBI, # years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (*CITE RELEVANT FIGS & TABLES* ).
This indicates that this is not a subtle effect.
However, the effect on L_PGS, r_max, and $\Delta DBH$, if any, is subtle, and a much longer record would be needed to detect such effects. 
Fortunately, we were able to get at annual growth using tree-ring data (Fig. `r tree_rings`).
These confirm that the effect of spring temperatures on annual growth is modest and generally not significant, while revealing a slightly more complex picture.
Although spring temperatures prior to May rarely had a significant influence on ...

## ("limitation in generalization, how to interpret/fix")
THe tree-rings also reveal that the strongest temperature sensitivity aligned with peak growing season.
This pattern is common across tree-ring studies in temperate deciduous forests.
(*1-2 sentences generalizing about climate sensitivity of temperate deciduous forest growth--Neil can do this most easily.*)
Thus, while our analysis is limited to two sites, we expect that the finding that spring temperatures have limited effect on annual tree growth will hold generally.

## ("contributions beyond")
Our findings contribute to a broader picture of the influence of spring temperatures on forest ecosystem function. 
Warm spring temperatures result in earlier leaf-out (REFS), an earlier start to net C sequestration (REFS), and--as shown here--earlier initiation of stem growth (*CITE RELEVANT FIGS & TABLES* ).
Yet, an earlier start to the growing season does not imply a longer growing season [Table `r hypothesis_table`, Figs `r pheno_Tsensitivity`, `r DOYtiming`;  @zani_increased_2020].
Rather, growth finishes sooner-- presumably because of *C sink limitation* [@zani_increased_2020].
If sink limitation controls the end of woody growing season,

## ("science is better now")
As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the increased carbon sequestration that is anticipated in most models [Table `r hypothesis_table`; Figs. `r pheno_Tsensitivity`, `r tree_rings`; @zani_increased_2020].
The implication is that temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that models are underestimating future atmospheric CO~2~ overall.

~~~~

>>>>

#Table numbers
hypothesis_table=1
table_species=2
mean_growth_parameters_table=3

@@

## some stats/ 
species_table <- read.csv(file = "tables_figures/species_table.csv")
n_dendro_trees <- sum(na.omit(species_table[, "n.trees"]))
n_tree_years <- sum(na.omit(species_table[, "n.tree.years"]))
n_cores <- sum(na.omit(species_table[, "n.cores"]))

@@

Metal dendrometer bands installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (Table `r table_species`). 
Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest, respectively. 
The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census (Table `r sample_size_by_year`) *(tags_per_year_preclean in data folder)*. 
Across years, the number of bands sampled averaged *~135* (range: 99-145) at SCBI and  *~717* (range: *700-755*) at Harvard Forest. 
In total, our analysis included `r n_tree_years` tree-years (Table `r table_species`).

```{r species, eval=TRUE, echo=FALSE, warning=FALSE}
knitr::include_graphics("tables_figures/species_table.png")
```

@@

### Tree-rings
Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
We analyzed records for the dominant ring- and diffuse- porous species at each site (Table `r table_species`), excluding species with other xylem architectures (*Juglans nigra* L. at SCBI, *Tsuga Canadensis* (L.) Carrière at Harvard Forest)
At SCBI, cores were collected as described in @helcoski_growing_2019. 
Briefly, cores were taken from living or recently dead trees $\ge$ 10 cm $DBH$, including a randomly sampled subset of all dominant species within the ForestGEO plot [live trees in 2010-2011; @bourg_initial_2013] and trees found newly dead in annual tree mortality in 2016 and 2017 [@helcoski_growing_2019].
At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
In total, we analyzed `r n_cores` cores from eleven species at SCBI and three species at Harvard Forest (Table `r table_species`).

@@

### Climate
For the dendrometer band analysis, we used climate data from local weather stations. 
Daily maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 [https://forestgeo.github.io/Climate/; @anderson-teixeira_forestgeoclimate_2020]. 
Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).
The R package climpact () was used to plot temperatures for visual inspection. 
Climpact also identifies readings that were greater than 3 standard deviations away from yearly means. These points were labeled as outliers and removed from the dataset. 
Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail). 
(*how were gaps filled for Harvard? Or was the record good?*)

For the tree-ring analysis, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020], again via the ForestGEO Climate Data Portal v1.0 [@anderson-teixeira_forestgeoclimate_2020].

@@

```{r climwin, out.width="80%", echo=FALSE, fig.align="center", fig.cap= climwin_caption}
# Created in RScripts/additional_figures.R
knitr::include_graphics("tables_figures/climwin_figure.png")
```

@@

```{r tree_rings, out.width="100%", echo=FALSE, fig.align="center", fig.cap= tree_rings_caption}
# Created in RScripts/additional_figures.R
knitr::include_graphics("tables_figures/tree_cores_tmax.tif")
```

@@

We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity...

Although the number of years analyzed here is limited (# years at SCBI, # years at HF), the signal is clear and consistent: for both species groups at both sites, warmer temperatures during virtually any part of the spring led to significantly earlier achievement of growth milestones (*CITE RELEVANT FIGS & TABLES* ).
This indicates that this is not a subtle effect.
However, the effect on L_PGS, r_max, and $\Delta DBH$, if any, is subtle, and a much longer record would be needed to detect such effects. 
Fortunately, we were able to get at annual growth using tree-ring data (Fig. `r tree_rings`).
These confirm that the effect of spring temperatures on annual growth is modest and generally not significant, while revealing a slightly more complex picture.
Although spring temperatures prior to May rarely had a significant influence on ...

THe tree-rings also reveal that the strongest temperature sensitivity aligned with peak growing season.
This pattern is common across tree-ring studies in temperate deciduous forests.
(*1-2 sentences generalizing about climate sensitivity of temperate deciduous forest growth--Neil can do this most easily.*)
Thus, while our analysis is limited to two sites, we expect that the finding that spring temperatures have limited effect on annual tree growth will hold generally.

Our findings contribute to a broader picture of the influence of spring temperatures on forest ecosystem function. 
Warm spring temperatures result in earlier leaf-out (REFS), an earlier start to net C sequestration (REFS), and--as shown here--earlier initiation of stem growth (*CITE RELEVANT FIGS & TABLES* ).
Yet, an earlier start to the growing season does not imply a longer growing season [Table `r hypothesis_table`, Figs `r pheno_Tsensitivity`, `r DOYtiming`;  @zani_increased_2020].
Rather, growth finishes sooner-- presumably because of *C sink limitation* [@zani_increased_2020].
As sink limitation controls the end of woody growing season [@zani_increased_2020], we might expect that an earlier start to growth buffers against unfavorable growth conditions during the peak growing season (e.g., drought).
*Do we see any evidence of this???*

As climate change accelerates and spring temperatures become increasingly warmer, growing seasons will start earlier, but this is unlikely to provide the increased carbon sequestration that is anticipated in most models [Table `r hypothesis_table`; Figs. `r pheno_Tsensitivity`, `r tree_rings`; @zani_increased_2020].
The implication is that temperate deciduous forests are likely to sequester less CO~2~ than most models suggest, implying that models are underestimating future atmospheric CO~2~ overall.

~~~~
## Cam's previous content
